### 2019-06-25 2-Chapters/1-Chapter/chapter1.tex
Lots of typos on Ch1 etc

% % Main problem
% \textbf{Main problems.}
%
At the root of the problems that motivate this thesis are the two questions of global warming, and the increase of the world population.
% https://en.wikipedia.org/wiki/History_of_radio
During the last 150 years, mankind has developed a wide range of different communication technologies, and since the late $1890$s wireless communications between manufactured devices have been made possible, and more and more frequent in our lives.
With the avent of Internet of Things networks (IoT), billions of autonomous low-power devices are expected to be deployed worldwide, for a large range of different applications.
It is now well-known that with the current trend of population increase and with the energy crisis, any newly deployed technology should be cheap and energy efficient,
as well as adapted to serve a large number of people and devices.

That is why we are interested in this thesis about the possible applications of embedding a certain kind of Machine Learning algorithms (Multi-Armed Bandit algorithms),
directly into the future IoT devices, in order to improve their battery life and reduce the energy cost of IoT networks, by optimizing the wireless communication thanks to embedded decentralized decision making.

@@

% Central broadcasting
Starting from FM radio music and news, then old black-and-white TV and color TV, the first systems were purely made for centralized broadcasting of information. In a large area, one antenna well located emits continuously in a fixed Radio Frequency (RF) band, and many devices can listen to this RF traffic and decode it, for instance to listen to music.
Different standards have been developed, and some of them are still in use today, and their nature requires a central authority that allocates RF bands to different usages (\eg, TV channels etc).
For example, in the city of Cesson-Sévigné where the campus of CentraleSupélec is located, a high RF tower emits on many standards, and in particular the \SI{92.3}{\mega\hertz} band is allocated to a French classical music radio (\emph{Radio Classique}).
Broadcast-based systems are not only used for TV or radio,
and for instance two other famous and worldwide deployed examples are the long-range global navigation satellite systems (such as USA's GPS or European Union's Galileo), and short-range infrared remote controllers.
% https://en.wikipedia.org/wiki/Satellite_navigation

% Two ways communications
However, this kind of mono-directional wireless communications were soon found to not be satisfactory for many applications, and thus bi-directional standards were defined from the $1910$s.
First developed by and for the military, and the navy (\eg, the Titanic sent the first SOS signal over radio in $1912$),
such applications have reached general public usage since the $1950$s.
The greatest achievement of this long-running R\&D domain was the avent of mobile telephony in the $1990$s, which has continued to be more and more present in people lives in the richer countries in the world, with the booming of cellphones since the $2000$s and smartphones since the $2010$s.
In most systems, there is still an antenna in charge of a large area (or cell), and many devices are able to receive and transmit data to the antenna.
Nowadays, the most well-known systems include WiFi, and systems deployed for mobile phone communications, with a long list of standards from 1G, 2G, 3G, 4G and now 5G, still in development in $2019$.
Even if both systems initially differed in usage, as 1G was designed to exchange only voice, and WiFi to link wireless devices at home to an Internet-connected box, they are conceptually very close.
As the systems are cellular, any device knows the antenna (or base station) it is associated with, and the bi-directional nature of their exchanges makes a centralized control possible.
In order to optimize the efficiency of the network, the base station is equipped with decision making algorithms that affects the devices in the cell to the available resources (time slots, RF bands, power etc).
To be able to communicate with the base station without interfering with each others, such affectation should be orthogonal, and led to systems based on Time Division Multiple Access (TDMA) and Frequency Division Multiple Access (FDMA), or Non-Orthogonal techniques only started to emerge recently (\eg, NOMA).
Traditionally, the base station aimed only at maximizing the Quality of Service (QoS), that englobes the data rate, latency, availability and other measures of performance of the wireless communications to and from the devices.
More recently, green radio has emerged as a solution to the tradeoff between maximizing QoS while minimizing the power consumption of base stations and devices.

% The opposite: decentralized broadcasting, and the age of IoT
Finally, a third kind of systems are of the opposite kind, and can be designed as decentralized broadcasting: an antenna is still in charge of many devices, but such devices can only send up-link packets, and the only down-link data they can receive are short acknowledgements sent from the base-station to indicate success or failure of every up-link packet.
This family of wireless systems are referred to as Internet of Things,
and a typical example of application is for sensor networks.

For the future development of smart grids, smart cities, smart homes, or smart agriculture, sensor networks are promised to be widely deployed.
Two examples of future applications that are already in deployment, in France or other countries, are connected buildings and connected agriculture.
For buildings, the main goal is to reduce the cost of heating empty buildings and use temperature sensor networks to get accurate and regular data about the temperature in each room and each floor, and let the centralized heat control optimize its cost and energy consumption.
For agriculture, one example can be to equipped every cow (in large farms) with a sensor that emits biological information, such as body temperature or stress level etc, in order to optimize the time of milking and to monitor the health of the animals.
%
% OK commencer par un paragraphe qui explique qu'on étudie les télécom dans le but de les rendre automatiquement plus efficace, plus verte, plus adaptative

This third kind of wireless systems are characterized by their decentralized nature: due to the really limited information that the base station can send to the devices, it is no longer possible to optimize the efficiency of the network from the centralized point-of-view of the base station.
In the present and future IoT networks, many devices of heterogeneous nature are using the same antenna for different applications.
A common hypothesis is that such IoT devices have a strong constraint on their power consumption, as most of them will be deployed without a direct power access and run on a battery, which duration should be maximized (typically more than 10 years).
Another common constraint for IoT devices is their low duty cycle, as most applications target one or a few messages to send every day, in strike opposition to the high data-rate pursued for centralized systems such as 4G/5G and WiFi.
%
Many different standards for IoT networks have been proposed,
and they consist in a specification for both the \emph{PHY}sical
and the \emph{M}edium \emph{A}ccess \emph{C}ontrol (\emph{MAC}) layer.
To quote some examples of standards, ZigBee, Z-Wave or Bluetooth are targeting for short-range communications, while LoRaWAN, SIGFOX, Ingenu or Weightless are designed for long-range communications.
We refer to the survey \cite{Centenaro16} for more details, and references in \cite{Azari18} or our recent article \cite{MoyBesson2019}.

@@

\textbf{The spectrum scarcity issue.}
%
A major problem of current wireless technologies is the issue of spectrum scarcity:
in most kinds of frequency bands, the entire RF spectrum is now allocated and free bands no longer exist, limiting the possibility of adding new usage.
As illustrated in Figure~\ref{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum} below, only a very small portion of the RF spectrum in the USA is not yet allocated, from $0$ to \SI{9}{\kilo\hertz}, while the rest of the displayed portion of the spectrum, from \SI{9}{\kilo\hertz} to \SI{300}{\giga\hertz}, is allocated to various usages, that goes from maritime radionavigation (historically the first usage of radio telecommunication) to space research, inter-satellite, mobile telephony and a large number of other applications.
%
% Traditional regulatory structures have been built for an analog model and are not optimized for cognitive radio.
Regulatory bodies in the world, like the \href{https://www.fcc.gov/}{Federal Communications Commission} in the USA (see \href{https://www.fcc.gov/}{\texttt{FCC.gov}})
or \href{http://www.tdf.fr/}{TDF} in France (see \href{https://www.TDF.fr/}{\texttt{TDF.fr}}, previously known as \emph{``TéléDiffusion de France''}), as well as different independent measurement campaigns, found that most radio frequency spectrum was inefficiently utilized,
meaning that while a band can be allocated to a certain unique usage, it can be free from any user in certain times and/or places.
We refer to \cite{patil2011survey} for a survey on worldwide spectrum utilization, and to \cite{valenta2010survey} for the situation in Europe.

Cellular network bands are overloaded in most parts of the world, but other frequency bands (such as military, amateur radio and paging frequencies) are insufficiently utilized.
Independent studies performed in some countries confirmed that observation, and concluded that spectrum usage highly depends on time and place.
Moreover, the fixed spectrum allocation prevents rarely used frequencies, like those assigned to specific services, from being used, even when any unlicensed users would not cause noticeable interference to the assigned service.
Therefore, in the last $15$ years, regulatory bodies in the world have been considering whether to allow unlicensed users in licensed bands if they would not cause any interference to licensed users.
These initiatives have initiated the research on \textbf{Cognitive Radio},
and especially on one subfield that we detail below, \textbf{Dynamic Spectrum Access (DSA)}.
%
%
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.65\textwidth,angle=90]{United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum.pdf}
    \caption[A chart representing the allocation of radio spectrum in the USA in $2016$]{A chart representing the allocation of radio spectrum in the USA in $2016$. Only a small share of the spectrum is not yet allocated, the top-left corner in white (it is in logarithmic scale from \SI{0}{\kilo\hertz} to \SI{300}{\giga\hertz}). \copyright{} Wikimedia.}
    % https://upload.wikimedia.org/wikipedia/commons/c/c7/United_States_Frequency_Allocations_Chart_2016_-_The_Radio_Spectrum.pdf?uselang=fr
    \label{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum}
\end{figure}

% \TODOL{Je peux améliorer (ou enlever) ce passage qui site J.Toledano, ou alors le commenter en plus de détails ? J'aime bien son rapport en fait (j'aurai du le lire au début de ma thèse !)}
% % https://www.economie.gouv.fr/files/files/PDF/french-spectrum-mission-executive-summary-2014-06-25.pdf
% % https://www.economie.gouv.fr/files/files/PDF/rapport-gestion-dynamique-spectre-2014-06-30.pdf
% To highlight that the spectrum scarcity is not only an issue in the USA,
% we quote the abstract of the ``Dynamic Spectrum Management For Innovation And Growth'' report by J. Toledano \cite{Toledano2014EnglishSummary,Toledano2014FrenchFull}, who was commissioned by the French government in $2014$.
% Different important problems are mentioned:
% %
% \begin{itemize}%\tightlist
%     \item
%     \emph{``In France, digital economy represents $5\%$ of GDP and affects $80\%$ of the French economy.''}
%     And the usage of wireless communications is nowadays worldwide and omnipresent:
%     \emph{``Radio frequencies are essential to many sectors: communications, broadcasting, transportation, satellite networks, energy networks and smart grids, public or private security, defense, etc.''}
%     Thus it is of highest interest to set up efficient, green, low-cost and durable new technologies for the future wireless standards, networks and devices.
%     %
%     \item
%     The radio spectrum is not only widely used for all kinds of applications, it is nowadays intensely used for some of them.
%     \emph{``Everyone agrees on the growing need for spectrum.
%     This growing need results from two phenomena.
%     On the one hand, mobile traffic should be multiplied by 13 to 25 fold between 2011 and 2017.
%     On the other hand, the development of new innovative services such as the Internet of Things and its multiple applications (smart cities, e-health...), could lead to a growing number of connected devices, up to 50 billion in 2020, according to estimations.''}
%     The first estimation was correct, and the mobile traffic is again \href{https://www.ericsson.com/en/mobility-report/future-mobile-data-usage-and-traffic-growth}{estimated} to be increased by 10 times between 2019 and 2025.
%     % https://www.ericsson.com/en/mobility-report/future-mobile-data-usage-and-traffic-growth
%     % https://www.cisco.com/c/en/us/solutions/collateral/service-provider/visual-networking-index-vni/white-paper-c11-738429.html
%     If the second estimation was overestimating the growth of the IoT market, forecasts predict  more than 64 billion IoT devices by 2025.
%     % https://www.businessinsider.com/internet-of-things-report?IR=T
%     \item
%     \emph{``Today, there are no more available frequencies in the easily exploitable frequency bands.
%     Moreover, it is getting more and more difficult to resort to classical methods of frequency bands liberation.''}
%     As illustrated above in Figure~\ref{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum}, and discussed in \cite{patil2011survey,valenta2010survey}.
%     \item
%     \emph{``Growing recourse to spectrum sharing, and in particular dynamic spectrum sharing, constitutes an important spectrum reserve.
%     A particular sharing form, the use of unlicensed bands, open to all and free, has been intensely developed with the growth of Wi-Fi use.''}
%     This thesis focusses on spectrum sharing in unlicensed bands, for IoT networks that lie at the opposite direction as the one underlying Wi-Fi networks: the spectrum sharing will be learned automatically by the end-devices, and not centrally decided by the base station.
% \end{itemize}

@@

% Il faut expliquer ce que c'est que les bandits, rapidement, et leur application à l'OSA
More precisely, in this thesis we focus on a certain kind of Reinforcement Learning models where the learner does not have access to the entire reaction of the environment after taking its action.
In other words, we focus on RL with limited feedback, when the player only sees the reward given by its action at every round, and not the reward that would have been given if she chose any other action.
This kind of limited feedback is called \emph{bandit information}, and we discuss the history and the concept of \emph{Multi-Armed Bandits} in details in the next Chapter~\ref{chapter:2}.
Clinical trials and treatment discovery were historically the first applications of MAB since $1930$s, with the early work of W. Thompson \cite{Thompson33}.
%
Multi-Armed Bandits (MAB) are a simple yet powerful example of the well-known exploration/exploitation dilemma:
when facing a set of $K$ actions whose effects on the environment are unknown, a learning must balance between
exploration of the unknown actions, in order to collect more information about them,
and exploitation of the best action according to its current knowledge.
%
The MAB problem have been studied in both the machine learning and the statistics community, since the $1950$s with pioneers like H. Robbins \cite{Robbins52}, and more recently it is an active field of research, since the late $1990$s \cite{Anantharam87a,Anantharam87b,auer1995gambling,Agrawal95}.
Research on MAB has produced a vast literature in the $2000$s \cite{Auer02,Auer02NonStochastic,Audibert2009minimax} and continues to be a topic of high interest, as illustrated by the surveys and books \cite{Bubeck12,LattimoreBanditAlgorithmsBook,Slivkins2019}, and the wide range of applications of MAB in the recent years.

\paragraph{From IT to Cognitive Radio.}
%
Similarly, we can narrow down the second field of study of this thesis, steps by steps:
from Information Technologies (IT), to telecommunications, to wireless communications, then to Software Defined Radio (SDR),
and finally to Cognitive Radio (CR).
%
The transition from the historical approach of hardware-based radio to SDR architectures and is a gradual process that started in the early $1990$s and has accelerated in the $2000$s.
% https://en.wikipedia.org/wiki/Software-defined_radio
A SDR is a radio communication system where components that have been traditionally implemented in hardware (\eg, mixers, filters, amplifiers, modulators/demodulators, detectors, etc) are instead implemented by means of software on a personal computer or embedded system, with the exception of a unique pair of receiver/transmitter antennas.
Even though the SDR paradigm was initiated by the USA defense research, over the past years the industry has started to be interested by SDR, and CR have got significant attention from the academia and industry as well.
% CR has been the first answer to solve spectrum scarcity issue which collected a huge attention in early 21st century.
%
CR is not a standard technology, and as such it does not have a single definition, so let us start by quoting the definition of two researchers whose works were the roots of the development of CR, in the last $20$ years.
% - technologies de l'information > télécommunications > communications sans fil > radio intelligente > DSA > OSA
%
% je peux faire comme Navik, donner des définitions de la CR, citant d'autres travaux
\begin{itemize}
    \item
    Joseph Mittola in $1999$ said that
    \emph{``a really smart radio that would be self-, RF- and user-aware, and that would include software technology and machine learning capabilities along with a lot of high-fidelity knowledge of the radio environment''} \cite{Mitola99}.

    \item
    Then Simon Haykin in $2005$ also said that
    \emph{``a CR is an intelligent wireless communication system that is capable of being aware of its surroundings, learning, and adapting its operating parameters (\eg, transmit power and carrier frequency) on the fly with an objective of providing reliable anytime, anywhere, and spectrally efficient communication''} \cite{Haykin05}.
    He was the first to propose to use CR for dynamic spectrum access.

@@

\paragraph{Opportunistic Spectrum Access.}
%
In OSA, the focus is on one SU accessing a licensed spectrum, occupied by PU that have a strict priority over the SU, which has to follow a ``listen-before-talk'' access scheme.
%
The following hypothesis are maid:
the SU is equipped with sensing,
and a consider a fixed and finite set of orthogonal channels, \ie, different frequency bands in a licensed spectrum.
For instance, it can be a set of three Wi-Fi channels at different frequency, emitted by the same Wi-Fi station.
Another hypothesis is that both PU and SU are synchronized in times, by sub-dividing the time in discrete time steps.
%
Thus if the SU spends a short time in the beginning of each time slot to do sensing, it can scan for the presence or absence of any PU before transmitting.
If a SU was able to sense for all the $K$ frequency bands, it could simply transmit in one of the free channels if any, or not transmit if all channels are used at a given time step.
However, sensing is known to be error-prone and costly, especially for wide-band sensing (for more details, see the surveys \cite{yucek2009survey,subhedar2011spectrum}), thus most works on OSA limit the sensing capacity of SU to sensing only one channel at a time. This assumption, along with the regulation that PU cannot be disturbed, enforces the SU to transmit in the channel that it sensed.
% non negligible cost of radio reconfiguration

Focussing on one SU in an OSA network, it has to sequentially decide a channel to sense (in the set of $K\geq2$ channels), then it performs sensing, and finally it transmits in this channel if it is free.
The goal of the SU is to minimize its energy consumption (we remind that we focus on green radio) and to maximize its uplink data rate, or equivalently, to maximize its number of successful transmission.
%
If the different channels are not uniformly used by the PU, and if we assume a stationary hypothesis on the PU traffic, then the goal of the SU boils down to explore the different channels and exploiting the best ones.
This frames the Opportunistic Spectrum Access problem as an exploration/exploitation problem, with a finite set of actions (the channels, also called \emph{arms}),
in a sequential action-then-feedback cycle (time steps are $t\in\N^*$),
under partial information feedback.
These three hypotheses are the ones that restrict from the general RL framework to the specific Multi-Armed Bandit case (see Figure~\ref{fig:1:ReinforcementLearningCycle}).

>>>>

% % Main problem
% \textbf{Main problems.}
%
At the root of the problems that motivate this thesis are the two questions of global warming, and the increase of the world population.
% https://en.wikipedia.org/wiki/History_of_radio
During the last 150 years, mankind has developed a wide range of different communication technologies, and since the late $1890$s wireless communications between man-made manufactured devices have been made possible, and more and more frequent in our lives.
With the avent of Internet of Things networks (IoT), billions of autonomous low-power devices are expected to be deployed worldwide, for a large range of different applications.
It is now a world-wide consensus that with the current trend of population increase and with the energy crisis, any newly deployed technology should be both cheap and energy efficient,
as well as adapted to serve a large number of people and devices.

That is why we are interested in this thesis about the possible applications of embedding a certain kind of Machine Learning algorithms (Multi-Armed Bandit algorithms),
directly into the future IoT devices, in order to improve their battery life and reduce the energy cost of IoT networks, by optimizing their wireless communications and learning to be self-organized, thanks to embedded decentralized decision making.

@@

% Central broadcasting
Starting from FM radio music and news, then old black-and-white TV and color TV, the first systems were purely made for centralized broadcasting of information. In a large area, one antenna well located emits continuously in a fixed Radio Frequency (RF) band, and many devices can listen to this RF traffic and decode it, for instance to listen to music.
Different standards have been developed, and some of them are still in use today, and their nature requires a central authority that allocates RF bands to different usages (\eg, TV channels etc).
For example, in the city of Cesson-Sévigné where my campus of CentraleSupélec is located, a high RF tower emits on many standards, and in particular the \SI{92.3}{\mega\hertz} band is allocated to a French classical music radio (\emph{Radio Classique}).
Broadcast-based systems are not only used for TV or radio,
and for instance two other famous and worldwide deployed examples are the long-range global navigation satellite systems (such as USA's GPS or European Union's Galileo), and short-range infrared remote controllers.
% https://en.wikipedia.org/wiki/Satellite_navigation

% Two ways communications
However, this kind of mono-directional wireless communications were soon found to not be satisfactory for many applications, and thus bi-directional standards were defined from the $1910$s.
First developed by and for the military and the navy (\eg, the Titanic sent the first SOS signal over radio in $1912$),
such applications have reached general public usage since the $1950$s.
The greatest achievement of this long-running R\&D domain was the avent of mobile telephony in the $1990$s, which has continued to be more and more present in people's lives in the richer countries in the world, with the booming of cellphones since the $2000$s and smartphones since the $2010$s.
In most systems, there is still a base station (with one or more antenna) in charge of a large area (or cell), and many devices are able to receive and transmit data to the base station.
Nowadays, the most well-known systems include WiFi, and systems deployed for mobile phone communications, with a long list of standards from 1G, 2G, 3G, 4G and now 5G, still in development in $2019$.
Even if both systems initially differed in usage, as 1G was designed to exchange only voice, and WiFi to link wireless devices at home to an Internet-connected box, they are conceptually very close.
As the systems are cellular, any device knows the antenna (or base station) it is associated with, and the bi-directional nature of their exchanges makes a centralized control possible.
In order to optimize the efficiency of the network, the base station is equipped with decision making algorithms that affects the devices in the cell to the available resources (time slots, RF bands, power etc).
To be able to communicate with the base station without interfering with each others, such affectation should be orthogonal, and led to systems based on Time Division Multiple Access (TDMA) and Frequency Division Multiple Access (FDMA), or Non-Orthogonal techniques only started to emerge recently (\eg, NOMA).
Traditionally, the base station aimed only at maximizing the Quality of Service (QoS), that englobes the data rate, latency, availability and other measures of performance of the wireless communications to and from the devices.
More recently, green radio has emerged as a solution to the tradeoff between maximizing QoS of the system, while minimizing the power consumption of base stations and devices.

% The opposite: decentralized broadcasting, and the age of IoT
Finally, a third kind of systems are of the opposite kind, and can be designed as decentralized broadcasting: an antenna is still in charge of many devices, but such devices can only send up-link packets, and the only down-link data they can receive are short acknowledgements sent from the base-station to indicate success or failure of every up-link packet.
This family of wireless systems are referred to as Internet of Things (IoT),
and a typical example of application is for sensor networks.

For the future development of smart grids, smart cities, smart homes, or smart agriculture, sensor networks are promised to be widely deployed.
Two examples of future applications that are already in deployment, in France or other countries, are connected buildings and connected agriculture.
For buildings, the main goal is to reduce the cost of heating empty buildings (at night etc) and use temperature sensor networks to get accurate and regular data about the temperature in all rooms and floors, and let the centralized heat control optimize its cost and energy consumption.
For agriculture, one example can be to equipped every cow (in large farms) with a sensor that emits biological information, such as body temperature or stress level etc, in order to optimize the time of milking and to monitor the health of the animals.
%
% OK commencer par un paragraphe qui explique qu'on étudie les télécom dans le but de les rendre automatiquement plus efficace, plus verte, plus adaptative

This third kind of wireless systems are characterized by their decentralized nature: due to the really limited information that the base station can send to the devices, it is no longer possible to optimize the efficiency of the network from the centralized point-of-view of the base station.
In the present and future IoT networks, many devices of heterogeneous natures are using the same antenna for different applications.
A common hypothesis is that such IoT devices have a strong constraint on their power consumption, as most of them will be deployed without a direct power access and run on a tiny battery, which lifespan should be maximized, as typically commercial IoT companies sell a lifespan more than 10 years \cite{Centenaro16}.
Another common constraint for IoT devices is their low duty cycles, as most applications target one or a few messages to send every day, in strike opposition to the high data-rate pursued for centralized systems such as 4G/5G and WiFi.
%
Many different standards for IoT networks have been proposed,
and they consist in a specification for both the \emph{PHY}sical
and the \emph{M}edium \emph{A}ccess \emph{C}ontrol (\emph{MAC}) layer.
To quote some examples of standards, ZigBee, Z-Wave or Bluetooth are targeting for short-range communications, while LoRaWAN, SIGFOX, Ingenu or Weightless are designed for long-range communications.
We refer to the survey \cite{Centenaro16} for more details, and references in \cite{Azari18} or our recent article \cite{MoyBesson2019}.

@@

\textbf{Spectrum Scarcity.}
%
A major problem of current wireless technologies is the issue of spectrum scarcity:
in most kinds of frequency bands, the entire RF spectrum is now allocated and free bands no longer exist, limiting the possibility of adding new usage.
As illustrated in Figure~\ref{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum} below, only a very small portion of the RF spectrum in the USA is not yet allocated, from $0$ to \SI{9}{\kilo\hertz}, while the rest of the displayed portion of the spectrum, from \SI{9}{\kilo\hertz} to \SI{300}{\giga\hertz}, is allocated to various usages, that goes from maritime radionavigation (historically the first usage of radio telecommunication) to space research, inter-satellite, mobile telephony and a large number of other applications.
%
% Traditional regulatory structures have been built for an analog model and are not optimized for cognitive radio.
Regulatory bodies in the world, like the \href{https://www.fcc.gov/}{Federal Communications Commission} in the USA (see \href{https://www.fcc.gov/}{\texttt{FCC.gov}})
or \href{http://www.tdf.fr/}{TDF} in France (see \href{https://www.TDF.fr/}{\texttt{TDF.fr}}, previously known as \emph{``TéléDiffusion de France''}), as well as different independent measurement campaigns, found that most radio frequency spectrum was inefficiently utilized,
meaning that while a band can be allocated to a certain unique usage, it can be free from any user in certain times and/or places.
We refer to \cite{patil2011survey} for a survey on worldwide spectrum utilization, and to \cite{valenta2010survey} for the situation in Europe.

Cellular network bands are overloaded in most parts of the world, but other frequency bands (such as military, amateur radio and paging frequencies) are insufficiently utilized.
Independent studies performed in some countries confirmed that observation, and concluded that spectrum usage highly depends on time and place.
Moreover, the fixed spectrum allocation prevents rarely used frequencies, like those assigned to specific services, from being used, even when any unlicensed users would not cause noticeable interference to the assigned service.
Therefore, in the last $15$ years, regulatory bodies in the world have been considering whether to allow unlicensed users in licensed bands, if they would not cause any interference to the (paying) licensed users.
These initiatives have initiated the research on \textbf{Cognitive Radio},
and especially on a field that we detail below, \textbf{Dynamic Spectrum Access (DSA)}.
%
%
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.65\textwidth,angle=90]{United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum.pdf}
    \caption[A chart representing the allocation of radio spectrum in the USA in $2016$]{A chart representing the allocation of radio spectrum in the USA in $2016$. Only a small share of the spectrum is not yet allocated, the top-left corner in white (it is in logarithmic scale from \SI{0}{\kilo\hertz} to \SI{300}{\giga\hertz}). \copyright{} Wikimedia.}
    % https://upload.wikimedia.org/wikipedia/commons/c/c7/United_States_Frequency_Allocations_Chart_2016_-_The_Radio_Spectrum.pdf?uselang=fr
    \label{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum}
\end{figure}

% \TODOL{Je peux améliorer (ou enlever) ce passage qui site J.Toledano, ou alors le commenter en plus de détails ? J'aime bien son rapport en fait (j'aurai du le lire au début de ma thèse !)}
% % https://www.economie.gouv.fr/files/files/PDF/french-spectrum-mission-executive-summary-2014-06-25.pdf
% % https://www.economie.gouv.fr/files/files/PDF/rapport-gestion-dynamique-spectre-2014-06-30.pdf
% To highlight that the spectrum scarcity is not only an issue in the USA,
% we quote the abstract of the ``Dynamic Spectrum Management For Innovation and Growth'' report by J. Toledano \cite{Toledano2014EnglishSummary,Toledano2014FrenchFull}, who was commissioned by the French government in $2014$.
% Different important problems are mentioned:
% %
% \begin{itemize}%\tightlist
%     \item
%     \emph{``In France, digital economy represents $5\%$ of GDP and affects $80\%$ of the French economy.''}
%     And the usage of wireless communications is nowadays worldwide and omnipresent:
%     \emph{``Radio frequencies are essential to many sectors: communications, broadcasting, transportation, satellite networks, energy networks and smart grids, public or private security, defense, etc.''}
%     Thus it is of highest interest to set up efficient, green, low-cost and durable new technologies for the future wireless standards, networks and devices.
%     %
%     \item
%     The radio spectrum is not only widely used for all kinds of applications, it is nowadays intensely used for some of them.
%     \emph{``Everyone agrees on the growing need for spectrum.
%     This growing need results from two phenomena.
%     On the one hand, mobile traffic should be multiplied by $13$ to $25$ fold between $2011$ and $2017$.
%     On the other hand, the development of new innovative services such as the Internet of Things and its multiple applications (smart cities, e-health...), could lead to a growing number of connected devices, up to $50$ billion in $2020$, according to estimations.''}
%     The first estimation was correct, and the mobile traffic is again \href{https://www.ericsson.com/en/mobility-report/future-mobile-data-usage-and-traffic-growth}{estimated} to be increased by $10$ times between $2019$ and $2025$.
%     % https://www.ericsson.com/en/mobility-report/future-mobile-data-usage-and-traffic-growth
%     % https://www.cisco.com/c/en/us/solutions/collateral/service-provider/visual-networking-index-vni/white-paper-c11-738429.html
%     If the second estimation was overestimating the growth of the IoT market, forecasts predict  more than 64 billion IoT devices by 2025.
%     % https://www.businessinsider.com/internet-of-things-report?IR=T
%     \item
%     \emph{``Today, there are no more available frequencies in the easily exploitable frequency bands.
%     Moreover, it is getting more and more difficult to resort to classical methods of frequency bands liberation.''}
%     As illustrated above in Figure~\ref{fig:1:United_States_Frequency_Allocations_Chart_2016_The_Radio_Spectrum}, and discussed in \cite{patil2011survey,valenta2010survey}.
%     \item
%     \emph{``Growing recourse to spectrum sharing, and in particular dynamic spectrum sharing, constitutes an important spectrum reserve.
%     A particular sharing form, the use of unlicensed bands, open to all and free, has been intensely developed with the growth of Wi-Fi use.''}
%     This thesis focusses on spectrum sharing in unlicensed bands, for IoT networks that lie at the opposite direction as the one underlying Wi-Fi networks: the spectrum sharing will be learned automatically by the end-devices, and not centrally decided by the base station.
% \end{itemize}

@@

% Il faut expliquer ce que c'est que les bandits, rapidement, et leur application à l'OSA
More precisely, in this thesis we focus on a certain kind of Reinforcement Learning models where the learner does not have access to the entire reaction of the environment after taking its action.
In other words, we focus on RL with limited feedback, when the player only sees the reward given by its action at every round, and not the reward that would have been given if she chose any other action.
This kind of limited feedback is called \emph{bandit information}, and we discuss the history and the concept of \emph{Multi-Armed Bandits} (MAB) in details in the next Chapter~\ref{chapter:2}.
Clinical trials and treatment discovery were historically the first applications of MAB since $1930$s, with the early work of W. Thompson \cite{Thompson33}.
%
MAB are a simple yet powerful example of the well-known exploration/exploitation dilemma:
when facing a set of $K$ actions whose effects on the environment are unknown, a learner must balance between
exploration of the unknown actions, in order to collect more information about them,
and exploitation of the best action, according to its current knowledge.
%
The MAB problem have been studied in both the machine learning and the statistics community, since the $1950$s with pioneers like H. Robbins \cite{Robbins52}, and more recently it is an active field of research, since the late $1990$s \cite{Anantharam87a,Anantharam87b,auer1995gambling,Agrawal95}.
Research on MAB has produced a vast literature in the $2000$s \cite{Auer02,Auer02NonStochastic,Audibert2009minimax} and continues to be a topic of high interest, as illustrated by the surveys and books \cite{Bubeck12,LattimoreBanditAlgorithmsBook,Slivkins2019}, and the wide range of applications of MAB in the recent years.

\paragraph{From IT to Cognitive Radio.}
%
Similarly, we can narrow down the second field of study of this thesis, steps by steps:
from Information Technologies (IT), to Telecommunications, to Wireless Communications, then to Software Defined Radio (SDR),
and finally to Cognitive Radio (CR).
%
The transition from the historical approach of hardware-based radio to SDR architectures and is a gradual process that started in the early $1990$s and has accelerated in the $2000$s.
% https://en.wikipedia.org/wiki/Software-defined_radio
A SDR is a radio communication system where components that have been traditionally implemented in hardware (\eg, mixers, filters, amplifiers, modulators/demodulators, detectors, etc) are all instead implemented by means of software on a personal computer or an embedded system, with the exception of a unique pair of receiver/transmitter antennas.
Even though the SDR paradigm was initiated by the USA defense research, over the past years the industry has started to be interested by SDR, and CR have got significant attention from the academia and industry as well.
% CR has been the first answer to solve spectrum scarcity issue which collected a huge attention in early 21st century.
%
CR is not a standard technology, and as such it does not have a single definition, so let us start by quoting the definition of two researchers whose works were the roots of the development of CR, in the last $20$ years.
% - technologies de l'information > télécommunications > communications sans fil > radio intelligente > DSA > OSA
%
% je peux faire comme Navik, donner des définitions de la CR, citant d'autres travaux
\begin{itemize}
    \item
    Joseph Mittola in $1999$ said that
    \emph{``a really smart radio that would be self-, RF- and user-aware, and that would include software technology and machine learning capabilities along with a lot of high-fidelity knowledge of the radio environment''} \cite{Mitola99}.

    \item
    Then Simon Haykin in $2005$ also said that
    \emph{``a CR is an intelligent wireless communication system that is capable of being aware of its surroundings, learning, and adapting its operating parameters (\eg, transmit power and carrier frequency) on the fly with an objective of providing reliable anytime, anywhere, and spectrally efficient communication''} \cite{Haykin05}.
    He was the first to propose to use CR for Dynamic Spectrum Access.

@@

\paragraph{Opportunistic Spectrum Access.}
%
In OSA, the focus is on one SU accessing a licensed spectrum, occupied by PU that have a strict priority over the SU, which has to follow a ``listen-before-talk'' access scheme.
%
The following hypothesis are maid:
the SU is equipped with sensing,
and a consider a fixed and finite set of orthogonal channels, \ie, different frequency bands in a licensed spectrum.
For instance, it can be a set of three Wi-Fi channels at different frequency, emitted by the same Wi-Fi station.
Another hypothesis is that both PU and SU are synchronized in times, by sub-dividing the time in discrete time steps.
%
Thus if the SU spends a short time in the beginning of each time slot to do sensing, it can scan for the presence or absence of any PU before transmitting.
If a SU was able to sense for all the $K$ frequency bands, it could simply transmit in one of the free channels if any, or not transmit if all channels are used at a given time step.
However, sensing is known to be error-prone and costly, especially for wide-band sensing, as detailed in the surveys \cite{yucek2009survey,subhedar2011spectrum}, thus most works on OSA limit the sensing capacity of SU to sensing only one channel at a time. This assumption, along with the regulation that PU cannot be disturbed, enforces the SU to transmit in the channel that it sensed, in each time step, if and only if it was sensed free from any PU.
% non negligible cost of radio reconfiguration

Focussing on one SU in an OSA network, it has to sequentially decide a channel to sense (in the set of $K\geq2$ channels), then it performs sensing, and finally it transmits in this channel if it was sensed free.
The goal of the SU is to minimize its energy consumption (we remind that we focus on green radio) and to maximize its uplink data rate, or equivalently, to maximize its number of successful transmission.
%
If the different channels are not uniformly used by the PU, and if we assume a stationary hypothesis on the PU traffic, then the goal of the SU boils down to explore the different channels and exploiting the best ones.
This frames the Opportunistic Spectrum Access problem as an exploration/exploitation problem, with a finite set of actions (the channels, also called \emph{arms}),
in a sequential action-then-feedback cycle (time steps are $t\in\N^*$),
under partial information feedback.
These three hypotheses are the ones that restrict from the general RL framework to the specific Multi-Armed Bandit case (see Figure~\ref{fig:1:ReinforcementLearningCycle}).
